---
title: 'Data Manipulation with R: Soil pore data'
author: "Edward Wallace"
date: "1/24/2018"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Why use R for data science?

We need to interpret and understand our data. To do this, we need to import, tidy, transform, visualize, and model data.

R makes this all *efficient*, *flexible*, and *reproducible*. With practice, you can not only make common things happen fast and knowing exactly what has happened, but you can access a wide variety of powerful tools to visualize and model your data. For free.

## Great resources on web and print

Data Carpentry: http://www.datacarpentry.org/R-ecology-lesson/

R for Data Science, Garrett Grolemund and Hadley Wickham: http://r4ds.had.co.nz/

Fundamentals of Data Visualization, Claus Wilke: http://serialmentor.com/dataviz/

Stack Overflow: https://stackoverflow.com/questions/tagged/r

Look for your local R meetup: https://www.meetup.com/LondonR/


# Description of Soil Pore dataset 

We (Amin Garbout) have CT-scanned the same soil, before rain (br), after rain (ar) and after a drying phase (ad).

We separated pores into 2 main categories:

  _cross : are pores that permit the drainages as they connect the soil surface to the below ground of the soil
  _unc :  are pores not connected to surface, they do not play a major role in water transport but have importance for soil structure
 

We measured key parameters from CT images:

  ID : pore id
  Vol : Volume (mm$^3$) of the pore
  x.Cent, y.Cent, z. Cent : centre of the pore in x, y, z co-ordinates (mm)
  Thickness, SD.Thickness, Max.Thickness : Mean, SD, Max thickness/ diameter of the pore (mm)
  SA : Surface area of the pore (mm) 


# Start: examples of base R functions

R can do arithmetic, and store many kinds of data: numbers, strings, and more.

```{r starting_examples}
1+1
my_number = 1000
my_subject = "stomata"

my_number
str(my_number)
str(my_subject)

```

## How to ask for Help

Importantly, you can ask for help using a question mark

```{r help}
?str

?base
```

Starting to type a command and pressing the tab key for suggestions is also a good idea.

# Many helpful collections of R functions are in packages

We're going to use the package "tidyverse"

```{r packages}
# install.packages("tidyverse")
library(tidyverse)
```

This loads  sub-packages for importing (readr), tidying (tibble), transforming (tidyr), and visualizing (ggplot2) data.

# Load data for before-rain cross.

We load the data with read_csv, which makes a "tibble". A tibble is a nicer version of the base-R "data frame" that holds many kinds of data in columns, where each row is one observation. This is like a more structured version of an excel table

```{r load_br_cross}
br_122_cross <- read_csv("data/br_122_cross.csv")
?read_csv
br_122_cross
```

## Examine it

Number of rows, structure, names, extract a single column.

```{r examine_br_cross}

nrow(br_122_cross)
str(br_122_cross)
names(br_122_cross)
br_122_cross$X1
```

Note that str gives too much information.

## Some of these columns look redundant, are they exactly the same?

```{r check_equal}
all.equal(br_122_cross$X1,br_122_cross$X.x)
all.equal(br_122_cross$X1,br_122_cross$X.x)
```

## Choose nicer column names and reload

We pick names. Then use set_names to set those. Last we "select" to remove the redundant columns.

```{r reload_br_cross}
pore_col_names <- c("ID","ID1","Source","ID2","Vol",
           "x.Cent","y.Cent","z.Cent",
           "Thickness","SD.Thickness","Max.Thickness","SA")

read_csv("data/br_122_cross.csv") %>%
    set_names(pore_col_names)

br_122_cross <- read_csv("data/br_122_cross.csv") %>%
    set_names(pore_col_names) %>%
    select("ID","Source","Vol","x.Cent","y.Cent","z.Cent",
           "Thickness","SD.Thickness","Max.Thickness","SA")
br_122_cross
```

## Summarize every column of the data

```{r summary_br_cross}
summary(br_122_cross)
```

# Tables are hard to understand: start visualizing

## Visualize x-and y-positions of pores

ggplot (for grammar of graphics plot) is a powerful structured way of plotting data. You need data in a tibble or data frame. Then you choose aesthetics, which are a way of mapping columns to plot features (position, size, colour, ...), and a geom which specifies how to do the mapping.

```{r xy_pos_br_cross}
ggplot(data=br_122_cross,aes(x=x.Cent,y=y.Cent)) +
    geom_point()
```

## Exercise: Visualize x- and z-positions

```{r xz_pos_br_cross,eval=FALSE,echo=FALSE}
ggplot(data=br_122_cross,aes(x=x.Cent,y=z.Cent)) +
    geom_point()
```

## Visualize Volume as size and and x- and y-position

```{r xy_pos_vol_br_cross}
ggplot(data=br_122_cross,aes(x=x.Cent,y=y.Cent,size=Vol)) +
    geom_point()
```


## Filter the data to consider outlying point separately

Filter is the verb to access a defined subset of the data.

```{r filter_y_cent_br_cross}
br_122_cross %>%
    filter(y.Cent < 9)

br_122_cross %>%
    filter(y.Cent > 9) 

ggplot(data=br_122_cross %>%
    filter(y.Cent > 9),
    aes(x=x.Cent,y=y.Cent,size=Vol)) +
    geom_point()

```

## Exercise: Load After Rain, cross, and visualize position-volume relation

From file "ar_122_cross.csv"

```{r load_ar_cross,eval=FALSE,echo=FALSE}
ar_122_cross <- read_csv("data/ar_122_cross.csv") %>%
    set_names(pore_col_names)
ar_122_cross
```

```{r xy_pos_vol_ar_cross,eval=FALSE,echo=FALSE}
ggplot(data=ar_122_cross,
    aes(x=x.Cent,y=y.Cent,size=Vol)) +
    geom_point()
```

# Put things you do again and again into functions

If we want to load multiple datasets in a very similar way, we can:

  - Copy and paste the same code with minor adjustments
  - Write a function

Writing functions is less error-prone and takes up less space. We've "chunked" - made a complicated task into a simple sub-task.

## Load pore data function

Mutate creates new or changed columns in a tibble. Here we use it to include the Soil and Pore category

```{r load_pore_data}
load_pore_data <- function(File,Soil,Cat) {
    read_csv(File) %>%
        set_names(pore_col_names) %>%
        mutate(Soil=Soil,Cat=Cat)
}

br_122_unc <- load_pore_data("data/br_122_unc.csv","br","unc")
```

## Now can we load all the soil conditions and pore categories?

Yes. With three pore types, and two pore categories, we can make a small tibble keeping track of all of those and the filenames (wiht mutate). 

```{r combinations}
soil_types <- c("br","ar","ad")
pore_cats <- c("cross","unc")

crossing(Soil=soil_types,Cat=pore_cats)

combinations <- crossing(Soil=soil_types,Cat=pore_cats) %>%
    mutate(File=paste0("data/",Soil,"_122_",Cat,".csv"))

combinations

```


# Use group_by and do to load all data

"group_by" tells our tibble we'd like it broken up a certain way.

"do" applies an arbitrary function to the output

Note: The syntax is a challenge. After a "pipe" (%>%) then the dot "." is used to refer to the thing, or grouped thing, that is fed in to the function.

```{r groups_do}
combinations %>%
    group_by(Cat,Soil)

pores_data <- 
    combinations %>%
    group_by(Cat,Soil) %>%
    do(load_pore_data(File=.$File,Cat=.$Cat,Soil=.$Soil))

```


## Summarize the data by group

Summarize is a simpler version of do for functions that return single values, not tibbles.

```{r summarize_pores_data}
groups(pores_data)

pores_data %>%
    group_by(Cat,Soil) %>%
    summarize(Vol.mean=mean(Vol),SA.mean=mean(SA),
              Thickness.mean=mean(Thickness))

```

Here we see that there are differences in thickness between soils. We will shortly return to visualize that.

## Write the cleaned, assembled data

Yes, there are functions for writing data, too, making it easier to return to.

```{r write_cleaned}
write_csv(pores_data,"data/pores_data_122.csv")
```

### Load data afresh from here if needed

```{r reread_cleaned,eval=FALSE}
pores_data <- read_csv("data/pores_data_122.csv")
```


# Visualize differences between soil conditions

## Histogram of Thickness for all cross pores

```{r Thickness_histogram}
ggplot(data=pores_data %>% 
           filter(Cat=="cross"),
       aes(x=Thickness)) +
      geom_histogram()
```

## Plot same data as a density curve

```{r Thickness_density}
ggplot(data=pores_data %>% 
           filter(Cat=="cross"),
       aes(x=Thickness)) +
      geom_density()
```


## Difference in pore thickness distribution by Soil

```{r Thickness_density_bySoil}
ggplot(data=pores_data %>% 
           filter(Cat=="cross"),
       aes(x=Thickness,colour=Soil)) +
      geom_density()
```

## Exercise: does Thickness vary with soil type for uncrossed pores?

```{r Thickness_density_bySoil_unc,eval=FALSE,echo=FALSE}
ggplot(data=pores_data %>% 
           filter(Cat=="unc"),
       aes(x=Thickness,colour=Soil)) +
      geom_density()
```

## Exercise: does that look better as a histogram?

```{r Thickness_density_bySoil_unc_limits,eval=FALSE,echo=FALSE}
ggplot(data=pores_data %>% 
           filter(Cat=="unc"),
       aes(x=Thickness,fill=Soil)) +
      geom_histogram() + 
    scale_x_continuous(limits=c(0,0.3))
```

# More on relationships between pairs of variables

## How are Volume and Surface Area related?

```{r Vol_SA}
ggplot(data=pores_data %>% 
           filter(Cat=="cross"),
       aes(x=SA,y=Vol)) +
    geom_point()
```

### Change to log-scales

```{r Vol_SA_loglog}
ggplot(data=pores_data %>% 
           filter(Cat=="cross"),
       aes(x=SA,y=Vol)) +
    geom_point() +
    scale_x_log10() + 
    scale_y_log10()
```


## Exercise: are Thickness and SA related by soil type?

```{r Vol_SA_loglog_soil,eval=FALSE,echo=FALSE}
ggplot(data=pores_data %>% 
           filter(Cat=="cross"),
       aes(x=Thickness,y=SA,colour=Soil)) +
    geom_point() +
    scale_x_log10() + 
    scale_y_log10()
```


## How could we summarize this relation?

introducing stat_smooth, a statistical summary of the data

```{r Vol_SA_loglog_smooth}
ggplot(data=pores_data %>% 
           filter(Cat=="cross"),
       aes(x=SA,y=Thickness,colour=Soil)) +
       geom_point() + 
    stat_smooth() +
    scale_x_log10() + 
    scale_y_log10() 
```

stat_smooth automatically uses a local smoothing method called loess.

## Summarize dependencies visually with a linear model

```{r Vol_SA_loglog_linear}
ggplot(data=pores_data %>% 
           filter(Cat=="cross"),
       aes(x=SA,y=Thickness,colour=Soil)) +
    geom_point() + 
    stat_smooth(method="lm",se=FALSE) +
    scale_x_log10() + 
    scale_y_log10() 
```

## Exercise: What about in uncrossed pores?

```{r Vol_SA_loglog_unc,eval=FALSE,echo=FALSE}
ggplot(data=pores_data %>% 
           filter(Cat=="unc"),
       aes(x=SA,y=Thickness,colour=Soil)) +
    geom_point() + 
    stat_smooth(method="lm",se=FALSE) +
    scale_x_log10() + 
    scale_y_log10() 
```


## Use facet_wrap to plot unc separately from crossed

facet_wrap allows breaking the data into separate panels

```{r Vol_SA_loglog_facet}
ggplot(data=pores_data,
       aes(x=SA,y=Thickness,colour=Soil)) +
    # geom_point() + 
    stat_smooth(method="lm",se=FALSE) +
    scale_x_log10() + 
    scale_y_log10() +
    facet_wrap(~Cat)
```

# What are the slopes of these lines? Use linear model(lm)

This is made hard by many steps involved.

## Sub-exercise: use a linear model to estimate relationship in one dataset

Note that the interesting correlation is with log10(SA); but in that case you have to remove the invalid values (log(0) = -Inf) to get the lm function to work.

```{r linear_model, eval=FALSE,echo=FALSE}
lm( data =  pores_data %>% 
        filter(Cat=="cross",Soil=="br",Thickness > 0, SA > 0 ),
 formula = log10(Thickness) ~ log10(SA))

```

## Use the broom package to tidy this output and wrap in a function

The broom package makes the outputs of linear models and so on into tidy tibbles.

```{r linear_model_function}
library(broom)

lm( data =  pores_data %>% 
        filter(Thickness > 0, SA > 0 ),
    formula = log10(Thickness) ~ log10(SA)) %>%
    tidy() 

lm_ThicknessSA <- function(sub_data) {
    lm( data =  sub_data %>% 
            filter(Thickness > 0, SA > 0 ),
        formula = log10(Thickness) ~ log10(SA)) %>%
    tidy() 
}

lm_ThicknessSA(pores_data)
```

### Sub-Exercise: use group_by and summarize to get the slopes of these lines.

```{r linear_model_function_group,echo=FALSE,eval=FALSE}
pores_data %>%
    group_by(Cat,Soil) %>%
    do(lm_ThicknessSA(.)) %>%
    filter(term=="log10(SA)") %>%
    select(Cat,Soil,estimate)
```


### Exercise: use "breaks" so that the scale labels are informative, and "theme_bw" to change the appearance of the plot

```{r informative_breaks,eval=FALSE,echo=FALSE}
ggplot(data=pores_data %>% 
           filter(Cat=="unc"),
       aes(x=SA,y=Thickness,colour=Soil)) +
    # geom_point() + 
    stat_smooth(method="lm",se=FALSE) +
    scale_x_log10() + 
    scale_y_log10(breaks=c(0.05,0.06,0.08,0.09,0.1,0.2,0.3,0.4,0.5)) +
    theme_bw() +
    theme(panel.grid.minor=element_blank())
```


## Some important topics we didn't cover

Other plots: lines (geom_line), boxplots (geom_boxplot), heatmaps (geom_tile), 2-d density plots (geom_hex), ...

Missing data: NA, na.omit, and so on.

Modeling: linear regression, clustering, etc.

All this and more can be done in R!

